---
title: "12_mefisto_full_model"
output: html_notebook
---

packages

```{r}
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("MOFAdata")
library(tidyverse)
library(MOFA2)
```


MEFISTO full model training 

```{r}
library(tidyverse)
library(MOFA2)

# ── 1. Load data ──────────────────────────────────────────────────────
soluble   <- read.csv("soluble_proteomics_clean_v2.csv")
cellular  <- read.csv("cellular_proteomics_clean_v2.csv")
insoluble <- read.csv("insoluble_proteomics_clean_v2.csv")

# ── 2. Standardise sample column names ───────────────────────────────
standardise_colnames <- function(df) {
  cols        <- names(df)
  sample_cols <- str_subset(cols, "^(D|M)\\d+\\..*\\.log2$")
  timepoints  <- str_extract(sample_cols, "^[DM]\\d+")
  new_names   <- ave(timepoints, timepoints, FUN = function(x) {
    paste0(x, "_rep", seq_along(x))
  })
  names(df)[match(sample_cols, names(df))] <- new_names
  df
}

cellular  <- standardise_colnames(cellular)
soluble   <- standardise_colnames(soluble)
insoluble <- standardise_colnames(insoluble)

# ── 3. Timepoint map (full six timepoints) ─────────────────────────────
timepoint_map <- c(D14 = 14, D21 = 21, M03 = 90, M06 = 180, M12 = 365, M18 = 548)

# ── 4. Build feature x sample matrices (no timepoint filtering) ───────
make_view_matrix <- function(df) {
  df %>%
    dplyr::select(gene_symbol, matches("^(D|M)\\d+_rep\\d+$")) %>%
    filter(!is.na(gene_symbol) & gene_symbol != "") %>%
    group_by(gene_symbol) %>%
    summarise(across(everything(), ~ mean(.x, na.rm = TRUE))) %>%
    ungroup() %>%
    column_to_rownames("gene_symbol") %>%
    as.matrix()
}

data_list_full <- list(
  cellular  = make_view_matrix(cellular),
  soluble   = make_view_matrix(soluble),
  insoluble = make_view_matrix(insoluble)
)

message("Cellular full:  ", nrow(data_list_full$cellular),  " proteins x ", ncol(data_list_full$cellular),  " samples")
message("Soluble full:   ", nrow(data_list_full$soluble),   " proteins x ", ncol(data_list_full$soluble),   " samples")
message("Insoluble full: ", nrow(data_list_full$insoluble), " proteins x ", ncol(data_list_full$insoluble), " samples")

# ── 5. Sample metadata (full timecourse) ──────────────────────────────
sample_cols_full <- colnames(data_list_full$cellular)

sample_meta_full <- tibble(sample = sample_cols_full) %>%
  mutate(
    timepoint_label = str_extract(sample, "^[DM]\\d+"),
    timepoint_num   = timepoint_map[timepoint_label],
    replicate       = str_extract(sample, "rep\\d+")
  )

message("Sample metadata:")
print(sample_meta_full)

# ── 6. Create MOFA object ──────────────────────────────────────────────
mofa_obj_full <- create_mofa(data_list_full)
samples_metadata(mofa_obj_full) <- sample_meta_full
mofa_obj_full <- set_covariates(mofa_obj_full, covariates = "timepoint_num")

# ── 7. Model options ───────────────────────────────────────────────────
model_opts_full             <- get_default_model_options(mofa_obj_full)
model_opts_full$num_factors <- 10

train_opts_full             <- get_default_training_options(mofa_obj_full)
train_opts_full$seed        <- 22
train_opts_full$convergence_mode <- "fast"

mefisto_opts_full           <- get_default_mefisto_options(mofa_obj_full)

mofa_obj_full <- prepare_mofa(
  mofa_obj_full,
  model_options    = model_opts_full,
  training_options = train_opts_full,
  mefisto_options  = mefisto_opts_full
)

# ── 8. Train model ─────────────────────────────────────────────────────
mofa_obj_full <- run_mofa(
  mofa_obj_full,
  outfile      = "mefisto_proteomics_full_timecourse.hdf5",
  use_basilisk = FALSE
)

message("Model trained and saved to mefisto_proteomics_full_timecourse.hdf5")

# ── 9. Extract outputs ─────────────────────────────────────────────────
factors_full <- get_factors(mofa_obj_full, as.data.frame = TRUE) %>%
  as_tibble() %>%
  left_join(sample_meta_full, by = "sample")

write_csv(factors_full, "mefisto_full_factor_scores.csv")
message("Saved factor scores")

weights_full <- get_weights(mofa_obj_full, as.data.frame = TRUE) %>%
  as_tibble() %>%
  mutate(
    feature     = as.character(feature),
    view        = as.character(view),
    factor      = as.character(factor),
    gene_symbol = case_when(
      str_ends(feature, "_cellular")  ~ str_sub(feature, 1, nchar(feature) - 9),
      str_ends(feature, "_soluble")   ~ str_sub(feature, 1, nchar(feature) - 8),
      str_ends(feature, "_insoluble") ~ str_sub(feature, 1, nchar(feature) - 10),
      TRUE ~ feature
    )
  )
write_csv(weights_full, "mefisto_full_weights.csv")
message("Saved protein weights")

var_exp_full <- calculate_variance_explained(mofa_obj_full)
var_df_full  <- var_exp_full$r2_per_factor[[1]] %>%
  as.data.frame() %>%
  rownames_to_column("factor") %>%
  pivot_longer(-factor, names_to = "view", values_to = "r2")

write_csv(var_df_full, "mefisto_full_variance_explained.csv")
message("Saved variance explained")

# ── 10. Significance of each factor vs timepoint (full timecourse) ────
factor_sig_full <- factors_full %>%
  group_by(factor) %>%
  summarise(
    p_value = tryCatch({
      fit <- lm(value ~ timepoint_num, data = cur_data())
      anova(fit)[["Pr(>F)"]][1]
    }, error = function(e) NA_real_),
    .groups = "drop"
  )

factor_sig_per_view_full <- var_df_full %>%
  left_join(factor_sig_full, by = "factor") %>%
  group_by(view) %>%
  mutate(padj = p.adjust(p_value, method = "BH")) %>%
  ungroup() %>%
  relocate(view, factor, r2, p_value, padj)

write_csv(factor_sig_per_view_full, "mefisto_full_factor_significance.csv")
message("Saved factor significance (p/padj) per fraction, full timecourse")

```{r}
library(tidyverse)

# ── Compute mean factor score per factor per timepoint ──────────────────
timepoint_levels <- names(timepoint_map)  # "D14" "D21" "M03" "M06" "M12" "M18"

factor_score_means <- factors_full %>%
  mutate(
    factor          = factor(factor, levels = paste0("Factor", 1:10)),
    timepoint_label = factor(timepoint_label, levels = timepoint_levels)
  ) %>%
  group_by(factor, timepoint_label) %>%
  summarise(mean_score = mean(value, na.rm = TRUE), .groups = "drop")

# ── Heatmap ───────────────────────────────────────────────────────────
p_factor_heatmap <- factor_score_means %>%
  mutate(factor = factor(factor, levels = paste0("Factor", 10:1))) %>%  # Factor1 at top
  ggplot(aes(x = timepoint_label, y = factor, fill = mean_score)) +
  geom_tile(colour = "black", linewidth = 0.3) +
  geom_text(aes(label = sprintf("%.2f", mean_score)),
            family = "Liberation Serif", size = 3, colour = "black") +
  scale_fill_gradient2(
    low      = "#2166ac",
    mid      = "white",
    high     = "#a50026",
    midpoint = 0,
    name     = "Mean\nfactor\nscore"
  ) +
  labs(
    title = "MEFISTO factor scores across the full ovarian ageing time course (D14\u2013M18)",
    x = "Timepoint",
    y = NULL
  ) +
  theme_minimal(base_family = "Liberation Serif", base_size = 13) +
  theme(
    plot.title       = element_text(size = 15, face = "bold", margin = margin(b = 10)),
    panel.grid       = element_blank(),
    axis.text.x      = element_text(size = 11),
    axis.text.y      = element_text(size = 12, face = "bold"),
    axis.title.x     = element_text(size = 14, margin = margin(t = 10)),
    legend.title     = element_text(size = 11)
  )

print(p_factor_heatmap)
ggsave("mefisto_full_factor_score_heatmap.pdf", p_factor_heatmap,
       width = 9, height = 6, device = cairo_pdf)
```



full model trajectories

```{r}

# Rebuild label data
timepoint_levels <- names(timepoint_map)  # "D14" "D21" "M03" "M06" "M12" "M18"

label_df <- factor_sig_full %>%
  mutate(
    factor    = factor(factor, levels = paste0("Factor", 1:10)),
    label_txt = paste0(
      "p = ", formatC(p_value, format = "e", digits = 2), "\n",
      "padj = ", formatC(padj, format = "f", digits = 3)
    ),
    x_pos = 1,
    y_pos = 1.8
  )

#  Rebuild plot_data
plot_data <- factors_full %>%
  mutate(
    factor          = factor(factor, levels = paste0("Factor", 1:10)),
    timepoint_label = factor(timepoint_label, levels = timepoint_levels)
  )

# Trajectory plot with title
p_trajectories_combined_smooth <- plot_data %>%
  ggplot(aes(x = as.numeric(timepoint_label), y = value)) +
  geom_smooth(method = "loess", se = TRUE, colour = "#0072B2", fill = "#0072B2",
              alpha = 0.15, linewidth = 1.3, span = 0.9) +
  geom_text(data = label_df,
            aes(x = x_pos, y = y_pos, label = label_txt),
            hjust = 0, vjust = 1, size = 3.2, family = "Liberation Serif",
            colour = "black", inherit.aes = FALSE, lineheight = 0.95) +
  facet_wrap(~ factor, nrow = 2, ncol = 5) +
  scale_x_continuous(breaks = seq_along(timepoint_levels),
                     labels = timepoint_levels,
                     expand = expansion(mult = c(0.1, 0.08))) +
  coord_cartesian(ylim = c(-2, 2)) +
  labs(
    title = "MEFISTO factor trajectories across the full ovarian ageing time course (D14-M18)",
    x = "Timepoint",
    y = "Factor value (fractions combined)"
  ) +
  theme_minimal(base_family = "Liberation Serif", base_size = 13) +
  theme(
    plot.title         = element_text(size = 16, face = "bold", margin = margin(b = 12)),
    panel.grid.minor   = element_blank(),
    strip.background   = element_rect(fill = "grey90", colour = NA),
    strip.text         = element_text(face = "bold", size = 12),
    axis.title.x       = element_text(size = 15, margin = margin(t = 12)),
    axis.title.y       = element_text(size = 15, margin = margin(r = 10)),
    axis.text.x        = element_text(angle = 90, hjust = 1, vjust = 0.5, size = 9),
    axis.text.y        = element_text(size = 10),
    panel.spacing      = unit(1, "lines"),
    plot.margin        = margin(10, 10, 15, 10)
  )

print(p_trajectories_combined_smooth)
ggsave("mefisto_full_factor_trajectories_combined_smooth_padj.pdf", p_trajectories_combined_smooth,
       width = 16, height = 8.5, device = cairo_pdf)
```

MEFISTO full model GO with reduced redundancy 

```{r}
library(rrvgo)
library(org.Mm.eg.db)
library(tidyverse)
# ── Load enrichment results ──────────────────────────────────────────
all_enrich <- read_csv("mefisto_full_model_enrichment/mefisto_full_all_factors_enrichment_by_fraction_SDsig.csv")

# ── Check what you have ──────────────────────────────────────────────
head(all_enrich)
names(all_enrich)

# ── Reduce GO annotation redundancy ──────────────────────────────────
reduce_go_terms_with_hierarchy <- function(df, ontology = "BP", threshold = 0.7) {
  go_df <- df %>% dplyr::filter(source == paste0("GO:", ontology))
  if (nrow(go_df) < 2) {
    message("  Too few GO:", ontology, " terms to reduce — skipping")
    return(go_df %>% dplyr::mutate(cluster_id = NA_integer_))
  }
  
  go_ids   <- go_df$term_id
  go_pvals <- setNames(go_df$p_value, go_ids)
  
  sim_matrix <- calculateSimMatrix(
    go_ids,
    orgdb    = "org.Mm.eg.db",
    ont      = ontology,
    method   = "Rel"
  )
  
  reduced <- reduceSimMatrix(
    sim_matrix,
    scores    = go_pvals,
    threshold = threshold,
    orgdb     = "org.Mm.eg.db"
  )
  
  # Convert to tibble and rename
  reduced_tib <- reduced %>%
    dplyr::as_tibble() %>%
    dplyr::rename(term_id = go) %>%
    dplyr::select(term_id, parentTerm, cluster)
  
  # Join back to get ALL terms with their cluster assignments
  result <- go_df %>%
    dplyr::left_join(reduced_tib, by = "term_id") %>%
    dplyr::mutate(
      cluster_id = cluster,
      is_representative = !is.na(parentTerm)
    )
  
  return(result)
}

# ── Apply per factor / fraction ──────────────────────────────────────
factors_list   <- paste0("Factor", 1:10)
fractions_list <- c("cellular", "soluble", "insoluble")
reduced_results <- list()

for (f in factors_list) {
  for (fr in fractions_list) {
    message("Now processing: ", f, " / ", fr)
    subset_df <- all_enrich %>% dplyr::filter(factor == f, label == fr)
    if (nrow(subset_df) == 0) next
    
    message("  Reducing GO:BP")
    bp_reduced <- reduce_go_terms_with_hierarchy(subset_df, ontology = "BP")
    message("  Reducing GO:MF")
    mf_reduced <- reduce_go_terms_with_hierarchy(subset_df, ontology = "MF")
    
    reduced_results[[paste(f, fr, "BP")]] <- bp_reduced
    reduced_results[[paste(f, fr, "MF")]] <- mf_reduced
  }
}

reduced_all <- dplyr::bind_rows(reduced_results)

# ── VIEW THE HIERARCHY: All terms grouped by cluster ──────────────────
hierarchy_view <- reduced_all %>%
  dplyr::filter(!is.na(cluster_id)) %>%
  dplyr::select(factor, label, source, cluster_id, parentTerm, term_name, is_representative, p_value) %>%
  dplyr::arrange(factor, label, source, cluster_id, is_representative, p_value)

print(hierarchy_view, n = 200)

# ── Summary: which terms group together ──────────────────────────────
cluster_summary <- reduced_all %>%
  dplyr::filter(!is.na(cluster_id)) %>%
  dplyr::group_by(factor, label, source, cluster_id, parentTerm) %>%
  dplyr::summarise(
    n_terms_in_cluster = n(),
    all_terms = paste(term_name, collapse = " | "),
    best_pvalue = min(p_value),
    .groups = "drop"
  ) %>%
  dplyr::arrange(factor, label, source, best_pvalue)

print(cluster_summary, n = 200)

# ── Save both views ──────────────────────────────────────────────────
write_csv(hierarchy_view, "GO_hierarchy_all_terms_per_cluster.csv")
write_csv(cluster_summary, "GO_clusters_summary.csv")
write_csv(reduced_all, "mefisto_full_GO_terms_reduced_with_hierarchy.csv")

message("✓ Saved: GO_hierarchy_all_terms_per_cluster.csv")
message("✓ Saved: GO_clusters_summary.csv")
message("✓ Saved: mefisto_full_GO_terms_reduced_with_hierarchy.csv")

#with gene symbol

library(org.Mm.eg.db)
library(AnnotationDbi)

# Read reduced GO terms
reduced <- read.csv("mefisto_full_model_enrichment/mefisto_full_GO_terms_reduced.csv")

# Get unique GO terms only
go_terms <- unique(reduced$term_id[grepl("^GO:", reduced$term_id)])

# Retrieve GO → gene mappings
go_gene_map <- AnnotationDbi::select(
  org.Mm.eg.db,
  keys = go_terms,
  keytype = "GO",
  columns = c("SYMBOL", "ENSEMBL")
)

# Collapse genes per GO term
go_genes_collapsed <- go_gene_map %>%
  group_by(GO) %>%
  summarise(
    gene_symbols = paste(sort(unique(na.omit(SYMBOL))), collapse = ";"),
    ensembl_ids  = paste(sort(unique(na.omit(ENSEMBL))), collapse = ";"),
    .groups = "drop"
  ) %>%
  rename(term_id = GO)

# Join back to reduced GO table
reduced_with_genes <- reduced %>%
  left_join(go_genes_collapsed, by = "term_id")

# Save
write.csv(
  reduced_with_genes,
  "mefisto_full_GO_terms_reduced_with_genes.csv",
  row.names = FALSE
)

library(ggplot2)
library(patchwork)

# LOAD DATA
kegg_data <- read.csv("mefisto_full_KEGG_terms.csv")
reac_data <- read.csv("mefisto_full_REAC_terms.csv")

factor_colours <- c("Factor1" = "#0072B2", "Factor2" = "#E69F00", "Factor3" = "#009E73")

# KEGG ENRICHMENT PLOT

kegg_factor_terms <- lapply(c("Factor1", "Factor2", "Factor3"), function(f) {
  df_terms <- kegg_data %>%
    filter(factor == f) %>%
    group_by(term_name) %>%
    summarise(
      p_value = min(p_value, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    mutate(factor = f)
  
  return(df_terms)
}) %>% bind_rows()

top_kegg_terms <- kegg_factor_terms %>%
  group_by(term_name) %>%
  summarise(min_p = min(p_value), .groups = "drop") %>%
  arrange(min_p) %>%
  slice_head(n = 15) %>%
  pull(term_name)

df_kegg_plot <- kegg_factor_terms %>%
  filter(term_name %in% top_kegg_terms) %>%
  mutate(
    term_name = factor(term_name, levels = rev(top_kegg_terms)),
    factor = factor(factor, levels = c("Factor1", "Factor2", "Factor3"))
  )

p_kegg <- ggplot(df_kegg_plot, aes(x = term_name, y = -log10(p_value), fill = factor)) +
  geom_col(position = "dodge", width = 0.7) +
  coord_flip() +
  scale_fill_manual(values = factor_colours, name = NULL) +
  labs(
    title = "KEGG pathway enrichment in top weighted proteins per MEFISTO factor",
    x = NULL,
    y = "-log10(p-value)"
  ) +
  theme_minimal() +
  theme(
    text = element_text(size = 9),
    axis.text.y = element_text(size = 8, colour = "black"),
    axis.text.x = element_text(size = 8, colour = "black"),
    axis.title.x = element_text(size = 10, colour = "black"),
    plot.title = element_text(size = 11, face = "bold", hjust = 0),
    panel.grid.major.x = element_line(colour = "grey90", linewidth = 0.3),
    panel.grid.minor = element_blank(),
    legend.position = "bottom",
    plot.margin = margin(t = 5, r = 5, b = 5, l = 5)
  )

ggsave(
  "mefisto_KEGG_enrichment_f123.pdf",
  p_kegg, width = 12, height = 8, dpi = 300, device = cairo_pdf
)

message("✓ KEGG plot saved: mefisto_KEGG_enrichment_f123.pdf")

# REACTOME ENRICHMENT PLOT
reac_factor_terms <- lapply(c("Factor1", "Factor2", "Factor3"), function(f) {
  df_terms <- reac_data %>%
    filter(factor == f) %>%
    group_by(term_name) %>%
    summarise(
      p_value = min(p_value, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    mutate(factor = f)
  
  return(df_terms)
}) %>% bind_rows()

top_reac_terms <- reac_factor_terms %>%
  group_by(term_name) %>%
  summarise(min_p = min(p_value), .groups = "drop") %>%
  arrange(min_p) %>%
  slice_head(n = 15) %>%
  pull(term_name)

df_reac_plot <- reac_factor_terms %>%
  filter(term_name %in% top_reac_terms) %>%
  mutate(
    term_name = factor(term_name, levels = rev(top_reac_terms)),
    factor = factor(factor, levels = c("Factor1", "Factor2", "Factor3"))
  )

p_reac <- ggplot(df_reac_plot, aes(x = term_name, y = -log10(p_value), fill = factor)) +
  geom_col(position = "dodge", width = 0.7) +
  coord_flip() +
  scale_fill_manual(values = factor_colours, name = NULL) +
  labs(
    title = "Reactome pathway enrichment in top weighted proteins per MEFISTO factor",
    x = NULL,
    y = "-log10(p-value)"
  ) +
  theme_minimal() +
  theme(
    text = element_text(size = 9),
    axis.text.y = element_text(size = 8, colour = "black"),
    axis.text.x = element_text(size = 8, colour = "black"),
    axis.title.x = element_text(size = 10, colour = "black"),
    plot.title = element_text(size = 11, face = "bold", hjust = 0),
    panel.grid.major.x = element_line(colour = "grey90", linewidth = 0.3),
    panel.grid.minor = element_blank(),
    legend.position = "bottom",
    plot.margin = margin(t = 5, r = 5, b = 5, l = 5)
  )

ggsave(
  "mefisto_REAC_enrichment_f123.pdf",
  p_reac, width = 12, height = 8, dpi = 300, device = cairo_pdf
)

message("✓ REAC plot saved: mefisto_REAC_enrichment_f123.pdf")

# COMBINED: GO + KEGG + REAC
# Re-run GO plot code from earlier
reduced_all <- read.csv("mefisto_full_GO_terms_reduced_with_genes.csv")
weights <- read.csv("mefisto_full_weights.csv")

all_factor_terms <- lapply(c("Factor1", "Factor2", "Factor3"), function(f) {
  top_proteins <- weights %>%
    filter(factor == f) %>%
    mutate(abs_value = abs(value)) %>%
    arrange(desc(abs_value)) %>%
    slice_head(n = 50) %>%
    pull(gene_symbol)
  
  df_terms <- reduced_all %>%
    filter(factor == f) %>%
    filter(source %in% c("GO:BP", "GO:MF")) %>%
    separate_rows(gene_symbols, sep = ";") %>%
    rename(gene_symbol = gene_symbols) %>%
    filter(gene_symbol %in% top_proteins) %>%
    group_by(parentTerm) %>%
    summarise(n_proteins = n_distinct(gene_symbol), .groups = "drop") %>%
    mutate(factor = f)
  
  return(df_terms)
}) %>% bind_rows()

top_terms <- all_factor_terms %>%
  group_by(parentTerm) %>%
  summarise(total = sum(n_proteins), .groups = "drop") %>%
  arrange(desc(total)) %>%
  slice_head(n = 15) %>%
  pull(parentTerm)

df_plot_bars <- all_factor_terms %>%
  filter(parentTerm %in% top_terms) %>%
  mutate(
    parentTerm = factor(parentTerm, levels = rev(top_terms)),
    factor = factor(factor, levels = c("Factor1", "Factor2", "Factor3"))
  )

p_go <- ggplot(df_plot_bars, aes(x = parentTerm, y = n_proteins, fill = factor)) +
  geom_col(position = "dodge", width = 0.7) +
  coord_flip() +
  scale_fill_manual(values = factor_colours, name = NULL) +
  labs(
    title = "GO term enrichment in top weighted proteins per MEFISTO factor",
    x = NULL,
    y = "Number of proteins"
  ) +
  theme_minimal() +
  theme(
    text = element_text(size = 9),
    axis.text.y = element_text(size = 8, colour = "black"),
    axis.text.x = element_text(size = 8, colour = "black"),
    axis.title.x = element_text(size = 10, colour = "black"),
    plot.title = element_text(size = 11, face = "bold", hjust = 0),
    panel.grid.major.x = element_line(colour = "grey90", linewidth = 0.3),
    panel.grid.minor = element_blank(),
    legend.position = "bottom",
    plot.margin = margin(t = 5, r = 5, b = 5, l = 5)
  )

message("✓ All enrichment plots created!")

#### SEX SPECIFIC TERMS/GENES

# Directionality of sex/reproductive GO terms
# Shows if genes in each term are upregulated (red) or downregulated (blue) 
# based on MEFISTO factor weights

library(tidyverse)
library(ggplot2)

# === Configuration ===
go_genes_file <- "mefisto_full_GO_terms_reduced_with_genes.csv"
weights_file <- "mefisto_full_weights.csv"
out_dir <- "."

# Sex/reproductive keywords
sex_keywords <- c(
  "sex", "sexual",
  "reproductive", "oogenesis", "spermatogenesis",
  "steroid", "hormone",
  "gonad", "ovarian", "testis",
  "folliculogenesis", "follicle",
  "ovulation"
)

# Okabe-Ito palette
oi_palette <- c(
  "Factor1" = "#0072B2",
  "Factor2" = "#E69F00",
  "Factor3" = "#009E73"
)

# === Load data ===
cat("Loading files...\n")
go_genes <- read_csv(file.path(out_dir, go_genes_file))
weights <- read_csv(file.path(out_dir, weights_file))

# === Extract sex/reproductive GO terms ===
keyword_pattern <- paste0("(?i)", paste(sex_keywords, collapse = "|"))

sex_go_data <- go_genes %>%
  filter(
    factor %in% c("Factor1", "Factor2", "Factor3"),
    str_detect(parentTerm, keyword_pattern)
  )

cat("Found", n_distinct(sex_go_data$parentTerm), "sex/reproductive GO terms\n")
cat("Found", nrow(sex_go_data), "term-factor combinations\n\n")

# === Extract individual genes from each term ===
sex_genes <- sex_go_data %>%
  # Split gene_symbols (semicolon-separated)
  separate_rows(gene_symbols, sep = ";") %>%
  rename(gene_symbol = gene_symbols) %>%
  select(parentTerm, gene_symbol, factor, p_value) %>%
  distinct() %>%
  # Join with MEFISTO weights (already in long format)
  left_join(
    weights %>% select(gene_symbol, factor, value) %>% rename(weight = value),
    by = c("gene_symbol" = "gene_symbol", "factor" = "factor")
  ) %>%
  filter(!is.na(weight))

cat("Found", n_distinct(sex_genes$gene_symbol), "unique genes in sex/reproductive terms\n\n")

# === Calculate mean weight per term per factor ===
term_directionality <- sex_genes %>%
  group_by(factor, parentTerm) %>%
  summarise(
    mean_weight = mean(weight, na.rm = TRUE),
    n_genes = n_distinct(gene_symbol),
    mean_pvalue = mean(p_value, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  # Clean up term names
  mutate(
    term_label = stringr::str_trunc(parentTerm, width = 60, side = "right"),
    factor = factor(factor, levels = c("Factor1", "Factor2", "Factor3"))
  ) %>%
  arrange(factor, mean_weight)

# === Create visualization ===
p <- ggplot(term_directionality, aes(x = reorder(term_label, mean_weight), 
                                      y = mean_weight,
                                      fill = ifelse(mean_weight > 0, "Upregulated", "Downregulated"))) +
  geom_col(width = 0.8) +
  facet_wrap(~factor, scales = "free", ncol = 1) +
  coord_flip() +
  scale_fill_manual(
    values = c("Upregulated" = "#d73027", "Downregulated" = "#4575b4"),
    name = "Direction"
  ) +
  geom_vline(xintercept = 0, colour = "black", size = 0.5) +
  labs(
    title = "Sex/Reproductive GO Terms: Gene Expression Direction by MEFISTO Factor",
    subtitle = "Red = upregulated (positive weights), Blue = downregulated (negative weights)",
    x = "Gene Ontology Term",
    y = "Mean MEFISTO Factor Weight"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 13, hjust = 0),
    plot.subtitle = element_text(size = 10, hjust = 0, colour = "grey40"),
    axis.title = element_text(face = "bold", size = 11),
    axis.text.x = element_text(size = 10, face = "bold"),
    axis.text.y = element_text(size = 11, face = "bold"),
    strip.text = element_text(face = "bold", size = 11),
    panel.grid.major.x = element_line(colour = "grey90", size = 0.3),
    panel.grid.minor.x = element_blank(),
    panel.grid.major.y = element_blank(),
    legend.position = "bottom"
  )

ggsave(file.path(out_dir, "go_sex_reproductive_directionality.pdf"), 
       plot = p, width = 14, height = 9, device = "pdf")
cat("Figure saved: go_sex_reproductive_directionality.pdf\n")

# === Export detailed summary ===
summary_detailed <- term_directionality %>%
  select(factor, term_label, mean_weight, n_genes) %>%
  rename(
    Factor = factor,
    `GO Term` = term_label,
    `Mean Weight` = mean_weight,
    `N Genes` = n_genes
  ) %>%
  arrange(Factor, `Mean Weight`)

write_csv(summary_detailed, file.path(out_dir, "go_sex_reproductive_directionality_summary.csv"))
cat("Summary table saved: go_sex_reproductive_directionality_summary.csv\n")

# === Print summary by factor ===
cat("\n=== Summary by Factor ===\n")
for (f in c("Factor1", "Factor2", "Factor3")) {
  upregulated <- sum(term_directionality$factor == f & term_directionality$mean_weight > 0)
  downregulated <- sum(term_directionality$factor == f & term_directionality$mean_weight < 0)
  cat(sprintf("%s: %d upregulated, %d downregulated\n", f, upregulated, downregulated))
}
```


MEFISTO subset M03-M12 model

```{r}
library(tidyverse)
library(MOFA2)

# 1. Load data
soluble   <- read.csv("soluble_proteomics_clean_v2.csv")
cellular  <- read.csv("cellular_proteomics_clean_v2.csv")
insoluble <- read.csv("insoluble_proteomics_clean_v2.csv")

# 2. Standardise sample column names
standardise_colnames <- function(df) {
  cols        <- names(df)
  sample_cols <- str_subset(cols, "^(D|M)\\d+\\..*\\.log2$")
  timepoints  <- str_extract(sample_cols, "^[DM]\\d+")
  new_names   <- ave(timepoints, timepoints, FUN = function(x) {
    paste0(x, "_rep", seq_along(x))
  })
  names(df)[match(sample_cols, names(df))] <- new_names
  df
}

cellular  <- standardise_colnames(cellular)
soluble   <- standardise_colnames(soluble)
insoluble <- standardise_colnames(insoluble)

# 3. Timepoint map
timepoint_map <- c(D14 = 14, D21 = 21, M03 = 90, M06 = 180, M12 = 365, M18 = 548)

# 4. Build feature x sample matrices
make_view_matrix <- function(df) {
  df %>%
    dplyr::select(gene_symbol, matches("^(D|M)\\d+_rep\\d+$")) %>%
    filter(!is.na(gene_symbol) & gene_symbol != "") %>%
    group_by(gene_symbol) %>%
    summarise(across(everything(), ~ mean(.x, na.rm = TRUE))) %>%
    ungroup() %>%
    column_to_rownames("gene_symbol") %>%
    as.matrix()
}

data_list <- list(
  cellular  = make_view_matrix(cellular),
  soluble   = make_view_matrix(soluble),
  insoluble = make_view_matrix(insoluble)
)

# 5. Filter to M03, M06, M12 only
keep_timepoints <- c("M03", "M06", "M12")

filter_to_timepoints <- function(mat, keep_tps) {
  keep_cols <- str_subset(
    colnames(mat),
    paste0("^(", paste(keep_tps, collapse = "|"), ")_rep\\d+$")
  )
  mat[, keep_cols, drop = FALSE]
}

data_list_subset <- list(
  cellular  = filter_to_timepoints(data_list$cellular,  keep_timepoints),
  soluble   = filter_to_timepoints(data_list$soluble,   keep_timepoints),
  insoluble = filter_to_timepoints(data_list$insoluble, keep_timepoints)
)

# Check dimensions
message("Cellular subset:  ", nrow(data_list_subset$cellular),
        " proteins x ", ncol(data_list_subset$cellular), " samples")
message("Soluble subset:   ", nrow(data_list_subset$soluble),
        " proteins x ", ncol(data_list_subset$soluble), " samples")
message("Insoluble subset: ", nrow(data_list_subset$insoluble),
        " proteins x ", ncol(data_list_subset$insoluble), " samples")

# 6. Sample metadata for subset
sample_cols_subset <- str_subset(
  colnames(data_list_subset$cellular),
  paste0("^(", paste(keep_timepoints, collapse = "|"), ")_rep\\d+$")
)

sample_meta_subset <- tibble(sample = sample_cols_subset) %>%
  mutate(
    timepoint_label = str_extract(sample, "^[DM]\\d+"),
    timepoint_num   = timepoint_map[timepoint_label],
    replicate       = str_extract(sample, "rep\\d+")
  )

message("Sample metadata:")
print(sample_meta_subset)

#7. Create MOFA object
mofa_obj_subset <- create_mofa(data_list_subset)
samples_metadata(mofa_obj_subset) <- sample_meta_subset
mofa_obj_subset <- set_covariates(mofa_obj_subset, covariates = "timepoint_num")

# 8. Model options
model_opts_sub             <- get_default_model_options(mofa_obj_subset)
model_opts_sub$num_factors <- 10

train_opts_sub             <- get_default_training_options(mofa_obj_subset)
train_opts_sub$seed        <- 22
train_opts_sub$convergence_mode <- "fast"

mefisto_opts_sub           <- get_default_mefisto_options(mofa_obj_subset)

mofa_obj_subset <- prepare_mofa(
  mofa_obj_subset,
  model_options    = model_opts_sub,
  training_options = train_opts_sub,
  mefisto_options  = mefisto_opts_sub
)

# ── 9. Train model
mofa_obj_subset <- run_mofa(
  mofa_obj_subset,
  outfile      = "mefisto_proteomics_M03_M06_M12.hdf5",
  use_basilisk = FALSE
)

message("Model trained and saved to mefisto_proteomics_M03_M06_M12.hdf5")

# ── 10. Extract outputs
# Factor scores
factors_subset <- get_factors(mofa_obj_subset, as.data.frame = TRUE) %>%
  as_tibble() %>%
  left_join(sample_meta_subset, by = "sample")

write_csv(factors_subset, "mefisto_M03_M12_factor_scores.csv")
message("Saved factor scores")

# Weights
weights_subset <- get_weights(mofa_obj_subset, as.data.frame = TRUE) %>%
  as_tibble() %>%
  mutate(
    feature     = as.character(feature),
    view        = as.character(view),
    factor      = as.character(factor),
    # Remove the view suffix that MOFA2 appends internally
    gene_symbol = case_when(
      str_ends(feature, "_cellular")  ~ str_sub(feature, 1, nchar(feature) - 9),
      str_ends(feature, "_soluble")   ~ str_sub(feature, 1, nchar(feature) - 8),
      str_ends(feature, "_insoluble") ~ str_sub(feature, 1, nchar(feature) - 10),
      TRUE ~ feature
    )
  )
write_csv(weights_subset, "mefisto_M03_M12_weights.csv")
message("Saved protein weights")

# Variance explained
var_exp_subset <- calculate_variance_explained(mofa_obj_subset)
var_df_subset  <- var_exp_subset$r2_per_factor[[1]] %>%
  as.data.frame() %>%
  rownames_to_column("factor") %>%
  pivot_longer(-factor, names_to = "view", values_to = "r2")

write_csv(var_df_subset, "mefisto_M03_M12_variance_explained.csv")
message("Saved variance explained")

#11a. Significance of each factor vs timepoint, per view 

factor_sig <- factors_subset %>%
  group_by(factor) %>%
  summarise(
    p_value = tryCatch({
      fit <- lm(value ~ timepoint_num, data = cur_data())
      anova(fit)[["Pr(>F)"]][1]
    }, error = function(e) NA_real_),
    .groups = "drop"
  )

# Join the same p-value onto every view for that factor (the test is on
# factor scores, which are shared across views by construction), then
# BH-adjust within each view's set of factors
factor_sig_per_view <- var_df_subset %>%
  left_join(factor_sig, by = "factor") %>%
  group_by(view) %>%
  mutate(padj = p.adjust(p_value, method = "BH")) %>%
  ungroup() %>%
  relocate(view, factor, r2, p_value, padj)

write_csv(factor_sig_per_view, "mefisto_M03_M12_factor_significance.csv")
message("Saved factor significance (p/padj) per fraction")
```


```{r}
library(tidyverse)

# Significance test for the subset model (already computed earlier
#    as factor_sig, but rebuilding here with padj to be safe
factor_sig_subset <- factors_subset %>%
  group_by(factor) %>%
  summarise(
    p_value = tryCatch({
      fit <- lm(value ~ timepoint_num, data = cur_data())
      anova(fit)[["Pr(>F)"]][1]
    }, error = function(e) NA_real_),
    .groups = "drop"
  ) %>%
  mutate(padj = p.adjust(p_value, method = "BH"))

write_csv(factor_sig_subset, "mefisto_M03_M12_factor_significance_combined.csv")
print(factor_sig_subset)

# Label data
timepoint_levels_subset <- c("M03", "M06", "M12")

label_df_subset <- factor_sig_subset %>%
  mutate(
    factor    = factor(factor, levels = paste0("Factor", 1:10)),
    label_txt = paste0(
      "p = ", formatC(p_value, format = "e", digits = 2), "\n",
      "padj = ", formatC(padj, format = "f", digits = 3)
    ),
    x_pos = 1,
    y_pos = 1.8
  )

# Plot data with discrete timepoint axis
plot_data_subset <- factors_subset %>%
  mutate(
    factor          = factor(factor, levels = paste0("Factor", 1:10)),
    timepoint_label = factor(timepoint_label, levels = timepoint_levels_subset)
  )

# Trajectory plot
p_trajectories_subset_smooth <- plot_data_subset %>%
  ggplot(aes(x = as.numeric(timepoint_label), y = value)) +
  geom_smooth(method = "loess", se = TRUE, colour = "#0072B2", fill = "#0072B2",
              alpha = 0.15, linewidth = 1.3, span = 0.9) +
  geom_text(data = label_df_subset,
            aes(x = x_pos, y = y_pos, label = label_txt),
            hjust = 0, vjust = 1, size = 3.2, family = "Liberation Serif",
            colour = "black", inherit.aes = FALSE, lineheight = 0.95) +
  facet_wrap(~ factor, nrow = 2, ncol = 5) +
  scale_x_continuous(breaks = seq_along(timepoint_levels_subset),
                     labels = timepoint_levels_subset,
                     expand = expansion(mult = c(0.15, 0.15))) +
  coord_cartesian(ylim = c(-2, 2)) +
  labs(x = "Timepoint", y = "Factor value (fractions combined)") +
  theme_minimal(base_family = "Liberation Serif", base_size = 13) +
  theme(
    panel.grid.minor  = element_blank(),
    strip.background  = element_rect(fill = "grey90", colour = NA),
    strip.text        = element_text(face = "bold", size = 12),
    axis.title.x       = element_text(size = 15, margin = margin(t = 12)),
    axis.title.y       = element_text(size = 15, margin = margin(r = 10)),
    axis.text.x        = element_text(angle = 0, hjust = 0.5, size = 10),
    axis.text.y        = element_text(size = 10),
    panel.spacing      = unit(1, "lines"),
    plot.margin        = margin(10, 10, 15, 10)
  )

print(p_trajectories_subset_smooth)
ggsave("mefisto_M03_M12_factor_trajectories_combined_smooth_padj.pdf", p_trajectories_subset_smooth,
       width = 16, height = 7.5, device = cairo_pdf)
```

integrated GO and scatterplot
```{r}
library(tidyverse)
library(patchwork)

# =====================================================================
# 0. Dissertation theme (consistent with your other figures)
# =====================================================================
dissertation_theme <- theme_minimal(base_size = 12, base_family = "Carlito") +
  theme(
    plot.title        = element_text(family = "Liberation Serif", face = "bold", size = 15, hjust = 0.5, colour = "black"),
    plot.subtitle     = element_text(family = "Carlito", face = "bold", size = 10, hjust = 0.5, colour = "grey30"),
    panel.background  = element_rect(fill = "white", colour = NA),
    panel.border      = element_rect(fill = NA, colour = "grey70", linewidth = 0.4),
    panel.grid.minor  = element_blank(),
    axis.title        = element_text(family = "Carlito", face = "bold", size = 11, colour = "black"),
    axis.text.x       = element_text(family = "Carlito", face = "bold", size = 9, colour = "black"),
    axis.text.y       = element_text(family = "Carlito", face = "bold", size = 9, colour = "black"),
    legend.position   = "none"
  )

category_colours <- c(
  "Concordant Up" = "#009E73",
  "Concordant Down" = "#E69F00",
  "Discordant (down RNA, up Protein)" = "#D55E00",
  "Discordant (up RNA, down Protein)" = "#0072B2"
)

# =====================================================================
# 1. Prep: top 10 terms per category, wrapped labels
# =====================================================================
plot_data <- enrichment_results %>%
  filter(source %in% c("GO:BP", "KEGG")) %>%
  group_by(category) %>%
  slice_min(order_by = p_value, n = 10, with_ties = FALSE) %>%
  ungroup() %>%
  mutate(
    neg_log10_p = -log10(p_value),
    term_label = str_wrap(term_name, width = 40)
  ) %>%
  arrange(category, neg_log10_p) %>%
  mutate(term_label = fct_inorder(term_label))

# =====================================================================
# 2. One barplot per category, coloured to match your scatter plot
# =====================================================================
make_category_barplot <- function(cat_name) {
  df <- plot_data %>% filter(category == cat_name)
  colour <- category_colours[[cat_name]]

  ggplot(df, aes(x = term_label, y = neg_log10_p)) +
    geom_col(fill = colour, width = 0.7) +
    coord_flip() +
    labs(
      title = cat_name,
      x = NULL, y = expression(-log[10](p[value]))
    ) +
    dissertation_theme
}

p_concordant_up   <- make_category_barplot("Concordant Up")
p_concordant_down <- make_category_barplot("Concordant Down")
p_discordant_dn_up <- make_category_barplot("Discordant (down RNA, up Protein)")
p_discordant_up_dn <- make_category_barplot("Discordant (up RNA, down Protein)")

p_concordant_up
p_concordant_down
p_discordant_dn_up
p_discordant_up_dn

# =====================================================================
# 3. Combine all four into one 2x2 patchwork figure
# =====================================================================
final_enrichment_figure <- (p_concordant_up | p_concordant_down) /
                            (p_discordant_dn_up | p_discordant_up_dn) +
  plot_annotation(
    title = "GO:BP and KEGG enrichment by RNA-protein integration category",
    subtitle = "Top 10 terms per category (FDR-corrected, custom background)",
    tag_levels = "A"
  ) &
  theme(
    plot.tag = element_text(family = "Liberation Serif", face = "bold", size = 14, colour = "black"),
    plot.title = element_text(family = "Liberation Serif", face = "bold", size = 17, hjust = 0.5, colour = "black"),
    plot.subtitle = element_text(family = "Carlito", face = "bold", size = 11, hjust = 0.5, colour = "grey30")
  )

final_enrichment_figure

# =====================================================================
# 4. Export
# =====================================================================
ggsave("integrated_rna_protein_enrichment_barplots.pdf", final_enrichment_figure,
       width = 16, height = 14, device = cairo_pdf)
ggsave("integrated_rna_protein_enrichment_barplots.png", final_enrichment_figure,
       width = 16, height = 14, dpi = 300, bg = "white")
```

```{r}
# ════════════════════════════════════════════════════════════════════
# PART C — SUMMARY STATISTICS
# ════════════════════════════════════════════════════════════════════

# ── C1. MEFISTO model summary ──────────────────────────────────────────
message("\n=== MEFISTO MODEL SUMMARY ===")

# Total variance explained per view (summed across all factors)
total_var_by_view <- var_df_subset %>%
  group_by(view) %>%
  summarise(total_r2 = sum(r2), .groups = "drop")
print(total_var_by_view)

# Number of factors that reach nominal / adjusted significance vs timepoint, per view
n_sig_factors <- factor_sig_per_view %>%
  group_by(view) %>%
  summarise(
    n_factors_total    = n(),
    n_sig_nominal_p05  = sum(p_value < 0.05, na.rm = TRUE),
    n_sig_padj05       = sum(padj < 0.05, na.rm = TRUE),
    .groups = "drop"
  )
message("\nSignificant factors per view (vs timepoint):")
print(n_sig_factors)

# Top variance-explaining factor per view
top_factor_per_view <- var_df_subset %>%
  group_by(view) %>%
  slice_max(r2, n = 1) %>%
  ungroup()
message("\nTop factor by variance explained, per view:")
print(top_factor_per_view)

# ── C2. Weight/loading distribution summary ────────────────────────────
message("\n=== LOADING DISTRIBUTION SUMMARY (per factor x view) ===")

loading_summary <- weights_z %>%
  group_by(factor, view) %>%
  summarise(
    n_proteins    = n(),
    mean_loading  = mean(value, na.rm = TRUE),
    sd_loading    = dplyr::first(sd_loading),
    n_z_gt1       = sum(abs(z_loading) > 1, na.rm = TRUE),
    n_z_gt2       = sum(abs(z_loading) > 2, na.rm = TRUE),
    .groups = "drop"
  )

# ── C3. DESeq2 summary ─────────────────────────────────────────────────
message("\n=== DESeq2 SUMMARY ===")

deseq_summary <- deseq %>%
  summarise(
    n_genes_total     = n(),
    n_up_nominal      = sum(log2FoldChange > 0 & pvalue < 0.05, na.rm = TRUE),
    n_down_nominal    = sum(log2FoldChange < 0 & pvalue < 0.05, na.rm = TRUE),
    n_up_padj         = sum(log2FoldChange > 0 & padj < 0.05, na.rm = TRUE),
    n_down_padj       = sum(log2FoldChange < 0 & padj < 0.05, na.rm = TRUE),
    median_abs_lfc    = median(abs(log2FoldChange), na.rm = TRUE)
  )
print(deseq_summary)

# ── C4. Integration summary ─────────────────────────────────────────────
message("\n=== INTEGRATION SUMMARY (relaxed filters, integrated_sig) ===")

integration_overview <- tibble(
  n_genes_raw_overlap   = length(overlap_genes),
  n_gene_factor_pairs_tested = nrow(integrated),
  n_gene_factor_pairs_sig    = nrow(integrated_sig),
  pct_concordant        = round(100 * mean(integrated_sig$direction %in% c("concordant_up", "concordant_down")), 1),
  pct_discordant        = round(100 * mean(integrated_sig$direction == "discordant"), 1)
)
print(integration_overview)

# Breakdown by fraction (view)
message("\nIntegration results by fraction:")
integrated_sig %>%
  group_by(view) %>%
  summarise(
    n_concordant_up   = sum(direction == "concordant_up"),
    n_concordant_down = sum(direction == "concordant_down"),
    n_discordant      = sum(direction == "discordant"),
    .groups = "drop"
  ) %>%
  print()

# Breakdown by factor (which factors contribute most to concordant vs discordant signal)
message("\nIntegration results by factor:")
integrated_sig %>%
  group_by(factor) %>%
  summarise(
    n_total       = n(),
    pct_discordant = round(100 * mean(direction == "discordant"), 1),
    .groups = "drop"
  ) %>%
  arrange(desc(n_total)) %>%
  print(n = 20)

# Save all summary tables to a single CSV bundle for easy reference
write_csv(loading_summary, "summary_loading_distribution.csv")
write_csv(deseq_summary, "summary_deseq2.csv")
write_csv(integration_overview, "summary_integration_overview.csv")
```

```{r}
# ── Summary table: concordant/discordant gene counts ──────────────────

# Table 1: counts by direction, collapsed across all factors/fractions
# (counts gene-factor-fraction combinations, not unique genes)
concordance_counts <- integrated_sig %>%
  count(direction) %>%
  mutate(pct = round(100 * n / sum(n), 1))

write_csv(concordance_counts, "concordance_counts_overall.csv")
print(concordance_counts)

# Table 2: counts by direction, broken down by fraction (view)
concordance_counts_by_view <- integrated_sig %>%
  count(view, direction) %>%
  group_by(view) %>%
  mutate(pct = round(100 * n / sum(n), 1)) %>%
  ungroup()

write_csv(concordance_counts_by_view, "concordance_counts_by_fraction.csv")
print(concordance_counts_by_view)

# Table 3: unique gene counts per direction (a gene counted once per
# direction it appears under, since it can differ across factors)
concordance_counts_unique_genes <- integrated_sig %>%
  distinct(gene_symbol, direction) %>%
  count(direction) %>%
  mutate(pct = round(100 * n / sum(n), 1))

write_csv(concordance_counts_unique_genes, "concordance_counts_unique_genes.csv")
print(concordance_counts_unique_genes)
```

```{r}

concordance_table_4cat <- integrated_sig %>%
  dplyr::count(direction_4cat) %>%
  dplyr::mutate(
    percentage = round(100 * n / sum(n), 1),
    direction_4cat = factor(direction_4cat, levels = c(
      "Concordant up (RNA up, protein up)",
      "Concordant down (RNA down, protein down)",
      "Discordant (RNA up, protein down)",
      "Discordant (RNA down, protein up)"
    ))
  ) %>%
  dplyr::arrange(direction_4cat) %>%
  dplyr::rename(Category = direction_4cat, `Number of genes` = n, `Percentage (%)` = percentage)

write_csv(concordance_table_4cat, "concordance_table_4category.csv")
print(concordance_table_4cat)
```

```{r}
# ── Reclassify ALL integrated genes into 5 categories (dissertation style) ──
integrated <- integrated %>%
  dplyr::mutate(
    direction_5cat = dplyr::case_when(
      !(sig_rna & sig_loading & passes_lfc)      ~ "Modest change",
      log2FoldChange > 0 & value > 0             ~ "Concordant Up",
      log2FoldChange < 0 & value < 0             ~ "Concordant Down",
      log2FoldChange < 0 & value > 0             ~ "Discordant (down RNA, up Protein)",
      log2FoldChange > 0 & value < 0             ~ "Discordant (up RNA, down Protein)",
      TRUE ~ NA_character_
    )
  )

# ── Summary table: N genes, % of total, mean RNA log2FC, mean protein weight ──
concordance_summary_table <- integrated %>%
  dplyr::group_by(direction_5cat) %>%
  dplyr::summarise(
    `N genes`          = dplyr::n(),
    `Mean RNA log2FC`  = round(mean(log2FoldChange, na.rm = TRUE), 2),
    `Mean protein weight` = round(mean(value, na.rm = TRUE), 2),
    .groups = "drop"
  ) %>%
  dplyr::mutate(
    `% of total` = round(100 * `N genes` / sum(`N genes`), 1),
    direction_5cat = factor(direction_5cat, levels = c(
      "Concordant Up",
      "Concordant Down",
      "Discordant (down RNA, up Protein)",
      "Discordant (up RNA, down Protein)",
      "Modest change"
    ))
  ) %>%
  dplyr::arrange(direction_5cat) %>%
  dplyr::rename(Category = direction_5cat) %>%
  dplyr::select(Category, `N genes`, `% of total`, `Mean RNA log2FC`, `Mean protein weight`)

write_csv(concordance_summary_table, "concordance_summary_table.csv")
print(concordance_summary_table)
```

```{r}
library(tidyverse)
library(ggplot2)
library(ggrepel)
integrated_unique_genes <- integrated_sig %>%
  dplyr::group_by(gene_symbol) %>%
  dplyr::slice_max(order_by = abs(z_loading), n = 1, with_ties = FALSE) %>%
  dplyr::ungroup()

# Build scatter data from your proteome-wide unique-gene table
# integrated_unique_genes already has one row per gene (strongest |z_loading|
# hit collapsed across factors/fractions), from the earlier proteome-wide step.
scatter_data <- integrated_unique_genes %>%
  filter(!is.na(padj)) %>%
  distinct(gene_symbol, .keep_all = TRUE) %>%
  mutate(category = case_when(
    log2FoldChange > 1  & value > 0 ~ "Concordant Up",
    log2FoldChange < -1 & value < 0 ~ "Concordant Down",
    log2FoldChange < -1 & value > 0 ~ "Discordant (down RNA, up Protein)",
    log2FoldChange > 1  & value < 0 ~ "Discordant (up RNA, down Protein)",
    TRUE ~ "Modest change"
  ))

#Top genes per category (by combined effect size) for labelling
top_genes <- scatter_data %>%
  filter(category != "Modest change") %>%
  group_by(category) %>%
  mutate(score = abs(log2FoldChange) * abs(value)) %>%
  arrange(desc(score)) %>%
  slice_head(n = 5) %>%
  ungroup()

# Plot
p_scatter <- ggplot() +
  # Modest change: fixed small size, plotted first so it sits behind the coloured points
  geom_point(data = scatter_data %>% filter(category == "Modest change"),
             aes(x = log2FoldChange, y = value),
             colour = "grey70", size = 1.2, alpha = 0.5) +
  # Concordant/discordant: size scaled by |protein weight|
  geom_point(data = scatter_data %>% filter(category != "Modest change"),
             aes(x = log2FoldChange, y = value, colour = category, size = abs(value)),
             alpha = 0.7) +
  scale_colour_manual(values = c(
    "Concordant Up" = "#009E73",
    "Concordant Down" = "#E69F00",
    "Discordant (down RNA, up Protein)" = "#D55E00",
    "Discordant (up RNA, down Protein)" = "#0072B2"
  ), name = NULL) +
  scale_size_continuous(range = c(1, 6), name = "|Protein weight|") +
  geom_text_repel(data = top_genes, aes(x = log2FoldChange, y = value, label = gene_symbol),
                  fontface = "bold", size = 4, max.overlaps = 100,
                  colour = "black", segment.colour = "grey50", segment.size = 0.3,
                  segment.alpha = 0.6) +
  labs(title = "RNA-Protein Integration Across the Proteome (M03-M12)",
       x = "RNA log2(Fold Change)", y = "Protein MEFISTO Weight") +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50", linewidth = 0.5) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey50", linewidth = 0.5) +
  theme_minimal(base_family = "Liberation Serif") +
  theme(plot.title = element_text(size = 12, face = "bold", hjust = 0),
        legend.position = "bottom",
        axis.text = element_text(size = 9))

cairo_pdf("integrated_rna_protein_scatter_proteome_wide_sized.pdf", width = 12, height = 8)
print(p_scatter)
dev.off()
```

```{r}
p_scatter <- ggplot() +
  # Modest change: fixed small size, plotted first so it sits behind the coloured points
  geom_point(data = scatter_data %>% filter(category == "Modest change"),
             aes(x = log2FoldChange, y = value),
             colour = "grey70", size = 1.2, alpha = 0.5) +
  # Concordant/discordant: size scaled by |protein weight|
  geom_point(data = scatter_data %>% filter(category != "Modest change"),
             aes(x = log2FoldChange, y = value, colour = category, size = abs(value)),
             alpha = 0.7) +
  scale_colour_manual(values = c(
    "Concordant Up" = "#009E73",
    "Concordant Down" = "#E69F00",
    "Discordant (down RNA, up Protein)" = "#D55E00",
    "Discordant (up RNA, down Protein)" = "#0072B2"
  ), name = NULL) +
  scale_size_continuous(range = c(1, 6), name = "|Protein weight|") +
  geom_text_repel(data = top_genes, aes(x = log2FoldChange, y = value, label = gene_symbol),
                  fontface = "bold", size = 4, max.overlaps = 100,
                  colour = "black", segment.colour = "grey50", segment.size = 0.3,
                  segment.alpha = 0.6) +
  labs(title = "RNA-Protein Integration Across the Proteome (M03-M12)",
       x = "RNA log2(Fold Change)", y = "Protein MEFISTO Weight") +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50", linewidth = 0.5) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey50", linewidth = 0.5) +
  theme_minimal(base_family = "Liberation Serif") +
  theme(plot.title = element_text(size = 12, face = "bold", hjust = 0),
        legend.position = "bottom",
        axis.text = element_text(size = 9))

cairo_pdf("integrated_rna_protein_scatter_proteome_wide_sized.pdf", width = 12, height = 8)
print(p_scatter)
dev.off()
```

